Instantiating Bunched Dependent Type Theory for Monoidal Classifying Toposes

### 2025-06-29 paper.tex
Bare bones argument for reverse direction. Needs formatting and reorganization.

\usepackage[style=alphabetic]{biblatex}
\addbibresource{paper.bib}

@@

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{construction}[theorem]{Construction}
\newtheorem{example}[theorem]{Example}

@@

@@

\printbibliography

>>>>

% \usepackage[style=alphabetic]{biblatex}
% \addbibresource{paper.bib}

@@

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{construction}[theorem]{Construction}
\newtheorem{example}[theorem]{Example}

@@

\section{Backwards}

TODO: Trying to make the main theorem an iff statement
TODO: This entire section will need to be rewritten once the proof works.
POSSIBILITY: Make the proof a chain of iffs. That way, there's a non-classifying characterization of topos models as well as a classifying one.
% culminating in a beautiful characterization in terms of classifying toposes.

Consider any model of \(\mathbf{BT}(*, 1, \Sigma, \Pi, \Pi^*)\) in the classifying topos over \(\T\).

TODO: Translate properties of the type theory to properties of the topos. E.g. \(*\) affine monoidal product...

Let \(U\) be the generic \(\T\)-model in \(\Set[\T]\).

\begin{lemma}
    For any \(A \in \Set[\T]\), the data \(A * -, \Pi^*_A\) is determined by the subset \(A * U \subseteq A \times U\).
\end{lemma}

\begin{proof}
    Observe that TODO form a geometric morphism \(\Set[\T]_{/A} \to \Set[\T]\).
    Each such geometric morphism yields a \(\T\)-model in \(\Set[\T]_{/A}\),
    given by the restriction of the obvious model \(\substack{A \times U \\ \downarrow \\ A}\) to \(\substack{A * U \\ \downarrow \\ A}\).
\end{proof}

\begin{corollary}
    The monoidal structure is determined by \(U * U\).
\end{corollary}

\begin{proof}
    Reduce from \(A*B\) to \(A*U\), then to \(U*U\).
\end{proof}

Next, we relate the classifying topos back to the syntactic category.

\begin{lemma}
    If an object of \(\Set[\T]\) is the interpretation of some context \(A \in \C_\T\) in the generic model,
    then every subobject is the interpretation of some restriction \((\vec{a} : A, \phi(\vec{a}))\).
    The choice of \(\phi\) is unique up to logical equivalence.
\end{lemma}

\begin{proof}
    Writing \(\Omega\) for the subobject classifier of \(\Set[\T]\), we have:
    \begin{align*}
        &\mathrm{Sub}(\mathbf{a}\yo A)
        \\ &= \mathbf{a}\yo A \to \Omega
        \\ &= \hom_{\Sh(\C_\T)}(\mathbf{a}\yo A, \Omega)
        \\ &= \hom_{\Psh(\C_\T)}(\yo A, i\Omega)
        \\ &= \Omega A
    \end{align*}
    This is the collection of covering-closed sieves on \(A\) in the syntactic category.

    A sieve has the required form when it contains exactly those morphisms that factor through the inclusion of \((\vec{a}:A,\phi(a))\) into \(A\).
    Uniqueness is immediate; if \(\phi_1\) and \(\phi_2\) are both solutions,
    then the inclusions of \((\vec{a}:A,\phi_1(a,b))\) and \((\vec{a}:A,\phi_2(a,b))\) factor through each other, producing a bi-implication.

    As for existince, fix any sieve \(S\).
    Every morphism it contains factors through the inclusion for the following choice of \(\phi\).
    \[\phi(\vec{a}) = \bigvee_{(B \in \C_\T, f : \hom_{\C_\T}(B, A)) \in S} \exists \vec{b} : B, f(\vec{b},\vec{a})\]
    And conversely, this inclusion is covered by the sieve, so covering-closedness ensures it it in the sieve.
\end{proof}

\begin{corollary}
    A geometric sequent \(\vec{a} : A \vdash \phi(\vec{a})\) is provable if it holds in the generic model.
\end{corollary}

\begin{proof}
    Uniqueness, from the above lemma, shows that \(\phi\) and \(\top\) are logically equivalent.
\end{proof}

\begin{corollary}
    A morphism in \(\Set[\T]\) is the interpretation of one in the generic model if its domain and codomain are.
\end{corollary}

\begin{proof}
    A morphism in \(\Set[\T]\) induces a provably functional relation.
    Both the relation and the proof of functionality are interpretations of the same in the syntactic category.
    This yields the data required to produce a morphism in \(\C_\T\).
\end{proof}

\begin{theorem}
    \(U * U\) is the interpretation of a unique geometrically-definable relation \((\sim) : X \times X \to \Prop\).
    This relation satisfies the assumptions of the main theorem.
\end{theorem}

\begin{proof}
    \(U * U\) is a subobject of \(U \times U\), which interprets \(X \times X\).
    So \(U \times U\) is the interpretation of some context \((a:X,b:X,a\sim b)\).

    It remains to show that \((\sim)\) is symmetric,
    and that \(\{a : X \mid a \sim x\}\) is a sub-\(\T\)-model of \(X\) for any \(x : X\).

    By the above lemma and its corollaries, it suffices to show that \((a,b) \in U*U \Leftrightarrow (b,a) \in (U*U)\),
    and that \(\{a : U \mid a \sim x\}\) is a sub-\(\T\)-model of \(U\) for any \(x : U\).

    The former is immediate. The latter holds because the geometric morphism TODO induces a \(\T\)-model in \(\Set[\T]_{/A}\).
\end{proof}

@@

% \printbibliography
